---
title: "Salary report by instructor level"
author: "Nick Cauldron"
date: "2024-02-04"
output:
  html_document:
    code_folding: hide
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data import

```{r, message=FALSE, warning=FALSE}
# Table parsing
# remotes::install_cran("tidyverse")
library(tidyverse)
library(scales)

# Adding gender
# remotes::install_cran("gender")
# remotes::install_github("lmullen/genderdata")
library(gender)
library(genderdata)
```

```{r}
salary_csv_osu_2023_f <- "salaries.csv"

salaries_f <- salary_csv_osu_2023_f
salaries_raw <- read_csv(salaries_f, show_col_types = FALSE) %>%
  janitor::clean_names()

salaries_raw
```

Convert the dates to real date columns to work with them.
For reference date we use the date the report was printed.
I found that by hand in the header on each page of the PDF.
This chunk also calculate some more info about how long the person has worked here.

Lastly it separates out names for gender prediction in the next chunk.
The first/middle name separation isn't perfect.
There are some edge cases where I can't tell if there are two first names or two middle names.
For the sake of simplicity, I use the first name following the comma as the "first name,"
and all after are part of their "middle name"

```{r clean_salary_report_dates}
report_print_date <- dmy("17-OCT-2023")
salaries <- salaries_raw %>%
  # clean organization column
  separate_wider_delim(job_orgn, delim = " - ",
                       names = c("job_orgn_code", "job_orgn_desc"),
                       cols_remove = FALSE, too_many = "merge") %>%
  # clean dates
  mutate(
    across(c(contains("date"), contains("hired")),
    dmy)) %>%
  mutate(appt_worked_days = report_print_date-appt_begin_date,
         rank_worked_days = report_print_date-rank_effective_date,
         appt_duration_days = appt_end_date-appt_begin_date,
         appt_remaining = appt_end_date-report_print_date) %>%
  mutate(appt_completed = case_when(appt_remaining > 0 ~ FALSE,
                                    appt_remaining <= 0 ~ TRUE,
                                    is.na(appt_remaining) ~ NA)) %>%
  # clean names - important for gender
  separate_wider_delim(name, delim = ", ",
                       names = c("name_last", "name_first_middle"),
                       cols_remove = FALSE) %>%
  separate_wider_delim(name_first_middle, delim = " ",
                       names = c("name_first", "name_middle"),
                       cols_remove = FALSE, too_few = "align_start", too_many = "merge") %>%
  # If someones first name is only one letter, use their middle name instead
  mutate(name_first = stringr::str_replace(name_first, "^\\w$", NA_character_)) %>%
  mutate(name_first = coalesce(name_first, name_middle))

glimpse(salaries)
```

Now add their expected gender.
[A person's age is important to accurately assign gender.](https://cran.r-project.org/web/packages/gender/vignettes/predicting-gender.html).
**I assumed an age range of**
**18-90 years old** (born 1934-2005).
We draw example names from Social Security Administration data in the United States from 1930-2012.

A more precise naming-by-age strategy could probably be designed for some jobs.
For example, there's a very high probability that assistant professors are younger than full professors.
It's probably not worth expanding on that for now.

```{r clean_salary_report_gender}
yr_birth_min <- 1934
yr_birth_max <- 2005
salaries_names_genders <- gender(unique(salaries$name_first), method = "ssa",
                                 years = c(yr_birth_min, yr_birth_max)) %>%
  select(-starts_with("year_"))
salaries_names_genders
```
```{r}
# join to salary date
# a few names were not in data,
# leave them as missing so models are easy and binary instead of a 3rd category "unknown"
salaries <- left_join(salaries, salaries_names_genders, by = c("name_first" = "name"))
```

Write to a file so others can join to salary data already parsed

```{r}
write_csv(salaries_names_genders, file = "names_genders_USA.csv")
```

Names assigned to gender based on simple majority rules.
Check confidence of that strategy by looking at the distribution of gender proportions that were assigned to each gender.

```{r}
salaries_names_genders_analysis <- salaries_names_genders %>%
  pivot_longer(cols = c("proportion_male", "proportion_female"),
               values_to = "proportion",
               names_to = c("gender"),
               names_repair = "unique") %>%
  rename("gender_assigned" = gender...2, "gender_proportions" = gender...3) %>%
  mutate(gender_proportions = str_remove(gender_proportions, "proportion_"))
salaries_names_genders_analysis
ggplot(salaries_names_genders_analysis, aes(x = gender_proportions, y = proportion)) +
  geom_jitter(width = 0.4, alpha = 0.2) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(vars(gender_assigned)) +
  labs(y = "Proportion of indivs of each gender",
       x = "Possible genders") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())
```

Widely, distribution of pay per gender?

```{r}
ggplot(salaries, aes(x = gender, y = annual_salary_rate)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.3, alpha = 0.2) +
  scale_y_log10(labels = scales::comma) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())
```